Article Title

### 2013-07-24 sections/Abstract.tex
expanded the theory section.

Linear correlations between the spiking activity of pairs of neurons are among the most familiar and useful descriptive statistics of neural activity. Multineuronal recordings provide far richer information than the equivalent number of neuron pairs considered in isolation. For example, the full covariance matrix reveals the partial correlations between pairs of neurons, as well as correlated activity across the entire population. Covariance matrix estimates require large samples for convergence. Convergence  can be improved by regularization, \emph{i.e.}~by imposing some kind of structure.  The optimal regularization must be determined empirically since its performance depends on how the imposed structure captures the underlying interactions. For example, we can use low-rank parameterizations of the covariance matrix to account for correlated input onto many neurons. Conversely, if correlations are a result of a small number of  pairwise linear interactions between the observed neurons, we can sparsify the inverse covariance matrix. We can also use a `sparse+low rank’ inverse covariance representation to account for common fluctuations and pairwise interactions in the recorded population.

>>>>

Correlations between the spiking activity of pairs of neurons are among the most familiar and useful descriptive statistics of neural activity. Multineuronal recordings provide richer information than the equivalent number of neuron pairs. For example, the full covariance matrix reveals correlated activity across the entire population as well as partial correlations between pairs.  Estimation of covariance matrices from finite recordings can be improved by regularization, \emph{i.e.}\;by imposing some kind of structure.  Optimal regularization must be determined empirically since its performance depends on how well the imposed structure captures the underlying regularities. For example, we can use low-rank parameterizations of the covariance matrix to account for common activity fluctuations across the recorded population. Conversely, if correlations are strongly influenced by a small fraction of possible pairwise linear associations between the observed neurons, we can impose sparsity on the inverse covariance matrix. We can also use a `sparse+low rank’ inverse covariance representation to account for both common fluctuations and pairwise interactions.

### 2013-07-24 sections/Introduction.tex
expanded the theory section.

\hl{\tiny forms of regularization} The convergence of the estimate of multivariate parameters to their true values can be improved through \emph{regularization},  Regularization is the deliberate biasing or \emph{shrinkage} of the empirical estimate toward a less variable \emph{target estimate} \citep{Schafer:2005} or the deliberate restriction of the estimate to a lower-dimensional space \citep{Dempster:1972,Fan:2008,Friedman:2008,Rothman:2008}.  A common intuitive explanation for the effect of regularization is that regularization adds some prior knowledge about the likely forms of the covariance matrix. Curiously, though, regularization need not reflect any plausible knowledge of the nature of covariance matrices in general or in the particular domain. Convergence can improve due to a purely statistical effect, sometimes known as \emph{Stein's phenomenon}, \emph{i.e.}~the favorable tradeoff between bias and variability, which arises \emph{anytime} a moderate bias toward a less variable target is mixed into an unbiased estimate.  Therefore, an estimator that is shown to converge faster on specific types of data does not necessarily contain within itself some wisdom of the statistical regularities in the data.  However, an estimator that does happen to accommodate the statistical structure of the data in a specific domain, it will gain additional advantage over an \emph{ad hoc} regularizer for that domain. Thus the optimal selection of the estimator for a specific problem with unknown statistical structure becomes an empirical question. And, if a clear winner emerges from a number of strong candidates, its regularization scheme can server as an effective statistical description of the covariance structure.

>>>>

\hl{\tiny forms of regularization} The convergence of the estimate of multivariate parameters to their true values can be improved through \emph{regularization},  Regularization is the deliberate biasing or \emph{shrinkage} of the empirical estimate toward a less variable \emph{target estimate} \citep{Schafer:2005} or the deliberate restriction of the estimate to a lower-dimensional space \citep{Dempster:1972,Fan:2008,Friedman:2008,Rothman:2008}.  A common intuitive explanation for the effect of regularization is that regularization adds some prior knowledge about the likely forms of the covariance matrix. Curiously, though, regularization need not reflect any plausible knowledge of the nature of covariance matrices in general or in the particular domain. Convergence can improve due to a purely statistical effect, sometimes known as \emph{Stein's phenomenon}, \emph{i.e.}\;the favorable tradeoff between bias and variability, which arises \emph{anytime} a moderate bias toward a less variable target is mixed into an unbiased estimate.  Therefore, an estimator that is shown to converge faster on specific types of data does not necessarily contain within itself some wisdom of the statistical regularities in the data.  However, an estimator that does happen to accommodate the statistical structure of the data in a specific domain, it will gain additional advantage over an \emph{ad hoc} regularizer for that domain. Thus the optimal selection of the estimator for a specific problem with unknown statistical structure becomes an empirical question. And, if a clear winner emerges from a number of strong candidates, its regularization scheme can server as an effective statistical description of the covariance structure.

### 2013-07-24 sections/Theory.tex
expanded the theory section.

\subsection*{Problem setup}
Let $x_1,\ldots,x_n \in \mathbb R^{p\times 1}$ denote a sample of $n$ observations of instantaneous activity (e.g.~spiking) rates of $p$ neurons averaged in time bins $i=1,\ldots,n$.

For simplicity, we assume that observations are sampled from a shared true distribution $x_1,\ldots,x_n \sim f$. 
\hl{\tiny explain: the derivations remain valid even without such a strong assumption.  We account for overdispersion due to temporal dependence and shuffle chunks of data to account for nonstationarities.}

For simplicity, we assume that the data have zero means.
We seek to estimate the true covariance matrix 
\begin{equation}
\Sigma=\mathbb E_{z\sim f}[ zz^\T]
\end{equation}

The purpose of the estimation procedure is to use the sample $x_1,\ldots,x_n$ to compute a covariance estimate $S$ that is as close as possible to $\Sigma$. 
The quality of the covariance estimate $S$ will be defined using a \emph{loss function} $\mathcal L\left(S, \Sigma\right)$, which is minimized when $S=\Sigma$.

In particular, we consider the mean-squared loss
\begin{equation}
\mathcal L_e( S, \Sigma) =  \| S - \Sigma\|^2_F 
\end{equation}
where $\| \cdot \|_F$ denotes the Frobenius norm;

and the multivariate normal log likelihood loss
\begin{equation}
\mathcal L_g(S,\Sigma) = \ln |S| +  \Tr(S^{-1} \Sigma)
\end{equation}
where $|S|$ denotes the determinant of matrix $S$.

\hl{\tiny list other possible loss functions and justify the choice of these two. Consider Stein's entropy loss and qudratic loss.   James:1961 showed failures of the MSE loss, but it continued to be widely used. Fan:2008 provides other arguments.}

The linearity of $\mathcal L_e(S,\Sigma)$ and $\mathcal L_g(S,\Sigma)$ with respect to the elements of $\Sigma$ allows unbiased empirical estimation of $\mathcal L$ from a validation sample $y_1,\ldots,y_m \sim f$ acquired independently of the original training sample $x_1,\ldots,x_n$ that was used to construct the estimate $S$.
\begin{equation}
\begin{split}
\mathcal L(S,\Sigma) & \equiv \mathbb E_{z\sim f} \left[ \mathcal L(S,zz^\T) \right] \\ 
&\approx \frac 1 m \sum\limits_{i=1}^m \mathcal L(S,y_i y_i^\T) = \mathcal L(S,\hat S_y)
\end{split}
\end{equation}
where $\hat S_y = \frac 1 m \sum\limits_{i=1}^m x_i x_i^\T$ is the empirical covariance of the validation dataset.

\hl{\tiny cross-validation}
In practice, neural recordings are limited.   Instead of validation with a dedicated validation dataset, $K$-fold cross-validation is used: The data are split into $K$ subsets of approximately equal size and each subset is used as the validation set with the covariance estimator trained on the remaning data.  The estimation of the loss from the $K$ tests is then averaged, which is justified, again, thanks to the linearity property of the chosen loss functions. 
\hl{\tiny exlain how the data are split into continuous chunks so that dependences between the training and validation datasets due to temporal correlations are minimized.  Coversely, each test set comprises multiple distirbuted chunks to reduce the effect of non-stationarities.}

>>>>

\subsection{Problem setup}
Let $x_1,\ldots,x_n$ denote a sample of observations of instantaneous spiking rates of $p$ neurons in time bins $i=1,\ldots,n$.  Each obseravation $x_i$ is a $p\times 1$ vector of spike counts.  We assume that observations independent, identically distributed, i.e.\;all observations are drawn from a common true distribution $x_1,\ldots,x_n \sim f$.  However, this assumptions can be relaxed and corrected for as discussed in this section. We make no assumption about the exact form of distribution $f$.

Then the true  covariance matrix $\Sigma$ is defined as
\begin{equation}
\Sigma =\mathbb E_z[ (z-\mathbb E_{z\sim f} [z])(z-\mathbb E_z [z])^\T]
\end{equation}
where $\mathbb E_z[g(z)]$ denotes the expected value of $g(z)$ when $z$ is distributed according to $f(z)$.

The purpose of the covariance estimation procedure is to obtain an estimate $\hat \Sigma$ that is as close as possible to $\Sigma$ given the finite sample  $x_1,\ldots,x_n$. We define a scalar \emph{loss function} $\mathcal L\left(\hat \Sigma, \Sigma\right)$ to measure the distance between $\hat \Sigma$ and $\Sigma$.  The loss function must have minimum at $\hat \Sigma=\Sigma$.

In particular, we consider the sum-of-squared-error loss
\begin{equation}
\mathcal L_e(\hat \Sigma, \Sigma) =  \| \hat \Sigma - \Sigma\|^2_F 
\end{equation}
where $\| A \|_F = \sqrt{\Tr(AA^\T)}$ denotes the Frobenius matrix norm; $\Tr(\cdot)$ denotes the matrix trace.

We also consider the normal-theory-log-likelihood loss
\begin{equation}
\mathcal L_g(\hat \Sigma,\Sigma) = \ln |\hat \Sigma|  + \Tr(\hat \Sigma^{-1}\Sigma)
\end{equation}
where $|A|$ denotes the determinant of matrix $A$.

Note that both $\mathcal L_e(\hat\Sigma,\Sigma)$ and $\mathcal L_g(\hat\Sigma,\Sigma)$ attain their minima when $\hat\Sigma = \Sigma$.   $\mathcal L_g$ is derived as the negative log likelihood function (scaled and shifted) when $f$ is a multivariate normal distribution.  However, using either of these loss functions does not amount to assuming a specific form of $f$. Other loss functions have been defended in statistical literature \citep{James:1961,Fan:2008}.

The loss function can be estimated based on another estimate $\hat\Sigma_Y$ of the covariance matrix obtained from an independent \emph{validation} sample $Y = (y_1,\ldots,y_m)\sim f^m$.

Both $\mathcal L_e(\hat\Sigma,\Sigma)$ and $\mathcal L_g(\hat\Sigma,\Sigma)$ are linear with respect to $\Sigma$ in the sense that
\begin{equation}
\mathcal L\left(\hat \Sigma,\alpha S_1 + (1-\alpha)S_2\right) 
\equiv 
\alpha\mathcal L(\hat\Sigma,S_1) + (1-\alpha)\mathcal L(\hat\Sigma,S_2)
\end{equation}
If we ensure that  $\hat\Sigma_Y$ is unbiased, i.e.\;$\mathbb E_Y\left[\hat \Sigma_Y\right]=\Sigma$, then this linearity implies that 
\begin{equation}
\mathbb E_Y \left[\mathcal L(\hat\Sigma,\hat\Sigma_Y)\right]
=
\mathcal L\left(\hat\Sigma,\mathbb E_Y \left[\hat\Sigma_Y\right]\right)
= 
\mathcal L(\hat\Sigma,\Sigma)
\end{equation}

In other words, the empirical loss $\mathcal L(\hat\Sigma,\hat\Sigma_Y)$ is an ubiased estimator of true loss $\mathcal L(\hat\Sigma,\Sigma)$ provided that $\hat\Sigma_Y$ is unbiased and $\hat\Sigma$ and $\hat\Sigma_Y$ are estimated from independent samples.  Note that $\hat\Sigma$ does not need to be unbiased.

In practice, recording a separate independent sample may not be sensible.  Instead, the recorded data a split into a training sample and validation sample.  One popular approach is $K$-fold validation, in which the data are split into $K$ roughly equal-sized subsets.  Each subset then successively serves as the validation  sample while the rest of the data make up the training sample. This procedure yields $k$ estimates of the loss, although not quite independent.
